---
title: "cmpt786_fp_random_forest"
output: pdf_document
date: "2025-03-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1 Setup

### 1.1 Libraries

Import the required libraries: 

```{r lib,echo=TRUE}
library(randomForest)
library(ranger)
```

### 1.2 Data Setup

Combined the malicious and normal datasets:

```{r data,echo=FALSE, comment=FALSE}
malicious <- read.csv("malicious_traffic.csv")
malicious$maliciousornot <- 1
normal <- read.csv("normal_traffic.csv")
normal$maliciousornot <- 0
all <- rbind(malicious, normal)

# Ensure maliciousornot is a factor
all$malicious <- as.factor(all$malicious)

# Use h (which has NAs removed) instead of all
if (!exists("h")) {
  h <- na.omit(all)
}

```

### 2.3 Train and Test Data Setup

```{r train&test,echo=FALSE, comment=FALSE}
# Set seed for reproducibility
set.seed(20876049)

# Define N and K
N <- nrow(h)
K <- 5

# Adjust N to be divisible by K (to avoid the earlier warning)
N_adjusted <- N - (N %% K)

# Create a matrix of indices for K-fold cross-validation
ind <- matrix(sample(1:N, size = N_adjusted), ncol = K)

# Define the column names (including the response variable)
namestt <- 
  c("StartTime", "Dur", "Proto", "SrcAddr", "Sport", "Dir", "DstAddr",  
    "Dport", "State", "sTos", "dTos", "TotPkts", "TotBytes", "SrcBytes", 
    "Label", "maliciousornot")

# Use the first fold as the test set and the remaining folds as the training set
test_indices <- ind[, 1]  # First fold for testing
train_indices <- unlist(as.list(ind[, -1]))  # Remaining folds for training

# Create the training and test sets
dtrain <- h[train_indices, ]
dtest <- h[test_indices, ]

# Set column names
colnames(dtrain) <- namestt
colnames(dtest) <- namestt

# Verify the sizes
cat("Training set size:", nrow(dtrain), "\n")
cat("Test set size:", nrow(dtest), "\n")

h <- na.omit(dtrain)
```

\newpage
## 2.1 Measuring the most suitable number of variables for the best model 

From the above, we could observe the lowest point is at 2 or 4, such that the most suitable number of variables for the best model is 2 or 4. Between the two numbers, since 2 is considerablely small and there will be limited information explained by 2 variables, we choose 4 as the most suitable number of variables for the best model.

```{r newranger}
# Set seed for reproducibility
set.seed(12345)

# Check the number of rows after removing NAs
N <- nrow(dtrain)
cat("Number of rows in h:", N, "\n")

# Split into training and test sets (2/3 train, 1/3 test)
N_train <- round(2 * N / 3)
N_test <- N - N_train
id.train <- sample(1:N, N_train, replace = FALSE)
id.test <- setdiff(1:N, id.train)

# Use ranger for better memory efficiency
library(ranger)

# Train random forest on a smaller sample first to test
# Take a sample of 10,000 rows from the training set
set.seed(12345)
train_sample <- h[id.train, ][sample(N_train, 10000), ]

# Run random forest with ranger
h.rf <- ranger(
  maliciousornot ~ StartTime + Dur + Proto + SrcAddr + Sport + Dir + 
    DstAddr + Dport + State + sTos + dTos + TotPkts + TotBytes + 
    SrcBytes + Label, data = train_sample, num.trees = 100, mtry = 7,
  importance = "impurity"
)

# Print the model
print(h.rf)

# Function to get explanatory variable names (fixed for ranger)
get.explanatory_varnames <- function(model) {
  return(model$forest$independent.variable.names)
}

# Prepare data for cross-validation
trainy <- as.factor(h[id.train, "maliciousornot"])  # Ensure factor for classification
trainx <- h[id.train, get.explanatory_varnames(h.rf)]

# Five-fold cross-validation on a sample to save memory
library(randomForest)
fb.rfcv <- rfcv(trainx = trainx[1:10000, ], trainy = trainy[1:10000], cv.fold = 5)

# Plot the results
with(fb.rfcv, plot(n.var, error.cv, pch = 19, type = "b", col = "blue"))

```

\newpage
## 2.2 Measuring the importance of Explanatory variables 

Due to the randomness, we run the following code for a few times. We need to choose the variables with the largest importance calculated. Based on the results, we could decide which variables have frequently appeared as the first four most important variables, that will be used for the following processes. The final decided variables are Label, Sport, SrcAddr and StartTime.

```{r imp,echo=TRUE}
### The type = 2 importance records the drop in RSS
importance(h.rf, type=2)
### The type = 1 importance records average decrease 
### in the accuracy of predictions
importance(h.rf, type=1)

### This is based on the average decrease in the accuracy of predictions 
### on “out of bag” samples when that variate is excluded from the model. 
### We will work primarily with the easier to comprehend type=1 measure of importance.
importance <- h.rf$variable.importance
importance_sorted <- sort(importance, decreasing = TRUE)
barplot(importance_sorted, 
        las = 2,  # Rotate x-axis labels for readability
        main = "Variable Importance (Ranger)", 
        ylab = "Importance")

### Running the code below with different set of seeds, 
### e.g. 123:543, 6789:7245, 20875800:20876099, 208758:208360,...
t2 <- rep(0,30)
t1 <- rep(0,30)
for (i in (6789:7245)){
  set.seed(i)
  h <- na.omit(all)
  N <- nrow(h)
  N_train <- round(2*N /3)
  N_test <- N - N_train
  # Select the two sample units
  id.train <- sample(1:N, N_train, replace=FALSE)
  id.test <- setdiff(1:N, id.train)
  # Run random forest with ranger
  h.rf <- ranger(
    maliciousornot ~ StartTime + Dur + Proto + SrcAddr + Sport + Dir + DstAddr + 
    Dport + State + sTos + dTos + TotPkts + TotBytes + SrcBytes + Label,
  data = train_sample, 
  num.trees = 100,  # Reduce number of trees to save memory
  mtry = 7,
  importance = "impurity"
)
  # The type = 2 importance records the drop in RSS
  i2 <- importance(h.rf, type=2)
  for (i in 1:30) {
    t2[i] <- t2[i] + i2[i]
  }
  i1 <- importance(h.rf, type=1)
  for (i in 1:30) {
    t1[i] <- t1[i] + i1[i]
  }
}
t1
t2
```
## 2.3 Error calculation algorithms used 
### 2.3.1 Functions for cross validation

```{r valid_train_functions, echo=TRUE}
# Set seed for reproducibility
set.seed(20876049)

# Define N as the number of rows in the dataset
N <- nrow(h)  # Assuming h is your dataset (from previous code)
K <- 5

# Adjust N to be divisible by K
N_adjusted <- N - (N %% K)

# Create a matrix of indices for K-fold cross-validation
ind <- matrix(sample(1:N, size = N_adjusted), ncol = K)

# Function to create the test set
testr <- function(data, indices) {
  indices <- indices[[1]][, 1]  # Extract the first column of indices
  te <- data[indices, ]  # Directly subset the data frame
  return(te)
}

# Function to create the training set
trainr <- function(data, indices) {
  indices <- indices[[1]]  # Extract the indices
  tr <- data[-indices, ]  # Exclude the test indices
  return(tr)
}

# Define the column names (predictors)
namestt <- 
  c("StartTime", "Dur", "Proto", "SrcAddr", "Sport", "Dir", "DstAddr",  
    "Dport", "State", "sTos", "dTos", "TotPkts", "TotBytes", "SrcBytes", 
    "Label", "maliciousornot")
```

\newpage
### 2.3.2 5-fold cross validation mean error

```{r newtrain}
# Initialize vector to store accuracy for each fold
accuracy <- c()
error <- c()
for (k in 1:5) {
  cur_ind <- list(subset(ind, select = k))
  train <- trainr(h, cur_ind)
  test <- testr(dtrain, cur_ind)
  colnames(train) <- namestt
  colnames(test) <- namestt

  # Sample the training data to avoid memory issues
  train <- train[sample(nrow(train), 10000), ]
  
  # Train the ranger model 
  h.bag <- ranger(
    maliciousornot ~ StartTime + Dur + Sport + Label, data = train,
    importance = "impurity"
    )
  
  # Predict on the test set
  pred <- predict(h.bag, data = test)$predictions 
  
  # Get actual values
  actual <- test$maliciousornot
  
  # Ensure consistent factor levels
  pred <- factor(pred, levels = c("0", "1"))
  actual <- factor(actual, levels = c("0", "1"))

  # Calculate accuracy for classification
  acc <- mean(pred == actual, na.rm = TRUE)

  # Calculate misclassification error = 1 - accuracy
  e <- 1 - acc
  error[k] <- e
}

# The mean misclassification error
mean(error)
```

### 2.3.3 Out Of Bag Error 

```{r oob, echo=TRUE}
h.bag <- ranger(
    maliciousornot ~ StartTime + Dur + Sport + Label,
  data = h)
pred <- h.bag$predictions
actual <- h$maliciousornot
e = sqrt(mean((pred-actual)^2))
e
```

The chosen model is the model built from random forest (or ranger), with 4 variables of StartTime, Dur, Sport and Label.

\newpage
## 2.4 Choosing mtry 
### for the best model built with 4 variables

The following code calculates the combination error of both 5-folds cross validation mean error and Out of Bag error for the best model built with 4 variables for mtry from 1 to 4.

For consideration of randomness, we run the code below for several times. The lowest combination error of both 5-folds cross validation mean error and Out Of Bag error is at mtry equal to 3, then we obtain the most suitable mtry is around 3. 

```{r mtry, echo=TRUE}
### 5-folds cross validation mean error for mtry from 1-4
merror <- c()
for (mt in 1:4){
  accuracy <- c()
  error <- c()
  for (k in 1:5) {
    cur_ind <- list(subset(ind, select = k))
    train <- trainr(h, cur_ind)
    test <- testr(dtrain, cur_ind)
    colnames(train) <- namestt
    colnames(test) <- namestt
    train <- train[sample(nrow(train), 10000), ]
    # Train the ranger model 
    h.bag <- ranger(
      maliciousornot ~ StartTime + Dur + Sport + Label, data = train,
      mtry = mt, importance = "impurity"
      )
    # Predict on the test set
    pred <- predict(h.bag, data = test)$predictions 
    # Get actual values
    actual <- test$maliciousornot 
    # Ensure consistent factor levels
    pred <- factor(pred, levels = c("0", "1"))
    actual <- factor(actual, levels = c("0", "1"))
    # Calculate accuracy for classification
    acc <- mean(pred == actual, na.rm = TRUE)
    # Calculate misclassification error = 1 - accuracy
    e <- 1 - acc
    error[k] <- e
  }
   merror[mt] <- mean(error)
}
merror

### Out Of Bag Error for ranges of mtry from 1-4
es <- c()
for (mt in 1:4){
  h.bag <- ranger(
      maliciousornot ~ StartTime + Dur + Sport + Label,
      data = h, mtry = mt)
  pred <- h.bag$predictions
  actual <- h$maliciousornot
  e = sqrt(mean((pred-actual)^2))
  es[mt] = e
}
es
combined_error = es + merror

bp <- barplot(
  combined_error,
  names.arg = c("1", "2", "3", "4"),
  xlab = "mtry Value",
  ylab = "Combined Error (OOB + Cross-Validation)",
  main = "Errors for the Chosen Random Forest Model \n with Different mtry Values",
  col = "steelblue",
  border = "black"
)
text(bp, combined_error, labels = round(combined_error, 4), pos = 3)
```
